\texttt{aeons}: approximating the end of nested sampling

### 2024-07-17 review_response/first_submission.tex
moderate revisions for journal submission

%{{{ Preamble
\documentclass[usenatbib]{mnras}
\usepackage{graphicx}
\usepackage[export]{adjustbox}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{subcaption}
\usepackage{layouts}
\usepackage{hyperref}
\usepackage[capitalise]{cleveref}
\usepackage{fancyvrb}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{gensymb}
\usepackage[T1]{fontenc}
%}}}

\renewcommand\thesubsubsection{\Alph{subsubsection}.}
\newcommand{\nlive}{n_i}
\newcommand{\Like}{\mathcal{L}}
\newcommand{\DKL}{\mathcal{D}_\mathrm{KL}}
\newcommand{\logLmax}{\log \Like_\mathrm{max}}
\newcommand{\dG}{d_\mathrm{G}}

@@

\section{Introduction}
Nested sampling is a multi-purpose algorithm invented by John Skilling which simultaneously functions as a probabilistic sampler, integrator and optimiser \citep{skilling}. It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics \citep{Trotta_2008}, materials science \citep{materials} and machine learning \citep{sparse_reconstruction}. The core algorithm is unique in its estimation of volumes by \textit{counting}, coupling together nested Monte Carlo integrals which makes high-dimensional integration feasible and robust. It also avoids problems that challenge traditional Bayesian samplers, such as posterior multi-modality and phase transitions.
\par
The order of magnitude run-time of an algorithm, that is, whether termination is hours or weeks and months away, is of high importance to the end user. Currently, existing implementations of nested sampling (e.g. \citealt{multinest, polychord, dnest, dynesty, ultranest, nessai,proxnest}) either do not give an indication of remaining run-time, or only provide crude measures of progress that do not directly correspond to the the true endpoint.
\par
This paper sets out a principled manner of endpoint estimation for nested sampling at each intermediate stage (as shown in \cref{fig:polychord_output}), the key idea being to use the existing samples to predict the likelihood in the region we have yet to sample from. We begin with an overview of nested sampling in \cref{sec:background}, followed by an examination of the anatomy of a nested sampling run to establish key concepts for endpoint prediction in \cref{sec:anatomy}. \cref{sec:endpoint} then outlines the methodology we use, including discussion and comparisons to previous attempts. Finally, \cref{sec:results} presents the results and discussions for toy and cosmological chains, before we conclude.
\begin{figure}
\begin{Verbatim}[frame=single, commandchars=\\\{\}]
\textcolor{red}{Predicted endpoint: 25054 +/- 242}
\textcolor{red}{Progress: [=================>########] 72%}
___________________
lives      |   500 |
phantoms   | 24310 |
posteriors | 18018 |
equals     |   245 |
-------------------
ncluster   =  1/1
ndead      =  18018
nposterior =  18018
nequals    =  249
nlike      =  4159049
<nlike>    =  491.04 (9.82 per slice)
log(Z)     =  -12.55 +/- 0.27
\end{Verbatim}
\caption{Output from \textsc{PolyChord} for a typical nested sampling run. The predicted endpoint, shown in red, is calculated using the method described in this paper.}
\label{fig:polychord_output}
\end{figure}

\section{Background}\label{sec:background}
We begin with a brief description of the nested sampling algorithm to establish the necessary notation. For a more comprehensive treatment, we recommend the original \citep{skilling} paper, the \citet{sivia} textbook, as well as the excellent technical review by \citet{Buchner_2023} and Nature review by \citet{physical_scientists}. 
\par
For a given likelihood $\mathcal{L}(\theta)$ and prior $\pi(\theta)$, nested sampling simultaneously calculates the Bayesian evidence
\begin{equation}\label{eq:evidence}
	\mathcal{Z} = \int \mathcal{L}\left(\theta\right)\pi(\theta)\ \mathrm{d}\theta
\end{equation}
while producing samples of the posterior distribution
\begin{equation}
	\mathcal{P}(\theta) = \frac{\mathcal{L}(\theta) \pi(\theta)}{\mathcal{Z}}.
\end{equation}
The algorithm operates by maintaining a set of $\nlive$ \textit{live points} sampled from the prior, which can vary in number throughout the run \citep{dynamic_ns}. At each iteration $i$, the point with the lowest likelihood is removed and added to a list of \textit{dead points} (illustrated in \cref{fig:dead_measure}). New points are then (optionally) drawn from the prior, subject to the constraint that they must have a higher likelihood than the current dead point~$\mathcal{L}_i$. Repeating the procedure leads to the live points compressing around peaks in the likelihood. If there are more births than deaths, the number of live points $n_i$ increases, whilst choosing to not generate new points reduces the number of live points. The exact creation schedule depends on the dynamic nested sampling strategy.
\par
The integral in \cref{eq:evidence} is then evaluated by transformation to a one-dimensional integral over the \textit{prior volume} $X$
\begin{equation}
	\mathcal{Z} = \int_0^1 \mathcal{L}(X)\ \mathrm{d}X \approx \sum_{i=1} \mathcal{L}_i\ \tfrac{1}{2}(X_{i-1}-X_{i+1}),
\end{equation}
where $X(\mathcal{L})$ is the fraction of the prior with a likelihood greater than $\mathcal{L}$. Whilst the likelihood contour $\mathcal{L}_i$ at each iteration is known, the prior volumes $X_i$ must be statistically estimated as follows: one can define a \textit{shrinkage factor} $t_i$ at each iteration $X_{i} = t_i X_{i-1}$, such that
\begin{equation}\label{eq:X_dist}
	X_i = \prod_{k=1}^i t_k.
\end{equation}
The $t_i$ are the maximum of $\nlive$ points drawn from $[0,1]$, so follow the distribution
\begin{equation}\label{eq:t_dist}
	P(t_i) = \nlive t_i^{\nlive-1},
\end{equation}
\begin{equation}\label{eq:t_moments}
    \langle\log t_i\rangle = -\frac{1}{\nlive}, \quad \mathrm{Var}(\log t_i) = \frac{1}{\nlive^2}.
\end{equation}
The algorithm terminates when an user-specified condition is met; a popular choice is when the evidence in the live points falls below some fraction $\epsilon$ of the accumulated evidence e.g. $10^{-3}$, which is proven to be a valid convergence criterion \citep{evans, Chopin_2010}.
Much of the existing literature treats this remaining evidence separately, for instance by estimating it as the termination $X$ multiplied by the average likelihood amongst the remaining live points. It is, however, quantitatively equivalent but qualitatively neater to consider termination as killing the remaining live points off one-by-one, incrementing the evidence exactly as during the run with decreasing $\nlive$ \citep{dynesty}.
\par
Uncertainties in the evidence are dominated by the spread in the prior volume distribution, and the simplest way to estimate them is by Monte Carlo sampling over sets of $\bm{t}$. As Skilling and others \citep{Chopin_2010, Keeton_2011} have shown, for any given problem the uncertainty in $\log \mathcal{Z}$ is proportional to $1/\sqrt{n_\mathrm{live}}$, so $n_\mathrm{live}$ sets the resolution of the algorithm.

@@

The bulk of the posterior lies within a prior volume ${X = e^{-\DKL}}$, which is the target compression. From \cref{eq:t_moments} one gets there by iteratively taking steps of size ${\Delta \log X_i = -1/n_i}$, so that when we add up the contribution of each step in \cref{eq:t_moments} we get

@@

\caption{The dead (small dots) and live points (larger dots) at several stages in a nested sampling run, recursively zoomed in. Note that the prior is taken to be uniform, which is reasonable because most nested sampling implementations require the prior to be provided as a transformation to the unit hypercube. The live points are uniform across the prior, while the dead points have a geometrically constant density.}

@@

The same is not true for the live points, whose enclosed prior volume values are uniformly distributed, and by the comments in the penultimate paragraph of \cref{sec:background} have ${n_{i+k} = n_i-k}$. As a result, the maximum live point is found at 

@@

The live points therefore only get us a factor of $\log \nlive$ closer in volume to the posterior bulk. In other words, it is not until we are around $\log \nlive$ away from $\log X = -\DKL$ that the samples begin populating the posterior typical set. One can see from \cref{eq:DKL} that the divergence increases linearly with dimension, so for large dimensionalities and typical live point numbers $\lesssim 1000$, this does not happen until near the end of the run.

@@

\subsection{Log-likelihood}\label{sec:logL}
We now consider the distribution of the samples in log-likelihood. To get an insight into the analytics, we will examine the case of the $d$-dimensional multivariate Gaussian likelihood with maximum point $\logLmax$ and length scale $\sigma$, situated in a uniform unit prior:
\begin{equation}\label{eq:logL_x}
    \log\Like = \logLmax - \frac{|\bm{x}|^{2}}{2\sigma^{2}}, \quad \bm{x} \in \mathbb{R}^{d}.
\end{equation}
This case is common and useful because many likelihoods are near-Gaussian, and most nested sampling implementations require the prior to be specified in terms of a transformation into the unit hypercube i.e. in prior probability units. Providing that $\sigma^d \ll 1$, the likelihood and posterior can be approximated as
\begin{equation}\label{eq:gaussian_logL}
    \log \Like = \logLmax - X^{2/d}/2\sigma^2,
\end{equation}
\begin{equation}
    \mathcal{P}(X) = \tfrac{1}{(2\pi\sigma^2)^{d/2}} e^{-X^{2/d}/2\sigma^2}, 
    \label{eqn:PX}
\end{equation}
which in terms of $\log\mathcal{L}$ becomes
\begin{equation}
    \mathcal{P}(\log\mathcal{L}) = \frac{1}{\Gamma\left(\frac{d}{2}\right)}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
\end{equation}
i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, with mean and variance
\begin{equation}
    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
\end{equation}
Given the prior volumes are uniform $X\sim U(0,1)$, the KL divergence in the $\sigma^d \ll 1$ limit is:
\begin{equation}
\mathcal{D}_\mathrm{KL} = -\frac{d}{2}\log 2 \Gamma(1+\tfrac{d}{2})^{2/d} \sigma^2.
\end{equation}

One can also derive the individual distributions for the live and dead points in log-likelihood. Note that this is merely the distribution of the $\log \Like$ \textit{values} of the points, without considering the associated prior volumes. The dead distribution is

@@

since the log-densities of the dead point prior volumes are proportional to the live point number at any given iteration. Meanwhile, the prior volumes enclosed by the live points (being independent samples) are always uniformly distributed, and so have the distribution 

@@

so that $y=0$ corresponds to the most recent dead point at $\log\Like_i$ and $y=1$ to the maximum. The live point distribution simplifies to

@@

which when analytically integrated to find its first and second moments give the mean and variance to be 

@@

     \pm \left( \frac{\Gamma(1+n_i)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n_i)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n_i)^2}{\Gamma(1+\frac{2}{d}+n_i)^2}\right)^{\frac{1}{2}}.

@@

In the large $d$, $n_i$ limit, taking the first and second order terms in the Taylor expansion simplifies the above expressions to 

@@

where $\psi^{(1)}$ is the trigamma function and $H_{n_i}$ is the $n_i$th harmonic number. The final term is a very small fraction in high dimensions, showing that for Gaussian and near-Gaussian cases the live points are far from the posterior bulk until the very end.

@@

    y_{(\nlive)}^{\mathrm{live}} \approx \frac{2\log \nlive}{d}.

@@

In other words, for a Gaussian the highest likelihood point is always at least an order of magnitude greater than the second highest. It is therefore typically the case that nearly all of the posterior mass is concentrated in a single point, the maximum live point, until the very end of the run when the prior volumes have shrunk enough to compensate.
The above results assume an exact Gaussian likelihood, which as discussed provides a reasonable first estimate for many physical likelihoods that are near-Gaussian. Further work can be done to generalise the analytics to non-Gaussians, though as we see in \cref{sec:results} results obtained via Gaussian assumptions can give the right order of magnitude.

@@

Previous literature \citep{Akrami_2010, Feroz_2011} has explored the potential for nested sampling to be used as a global maximiser, given its ability to handle multi-modalities. In particular, the latter authors emphasised that posterior samplers such as nested sampling find the bulk of the \textit{mass}, not the maximum of the distribution, but that this can be remedied by tightening the termination criterion. We now use the machinery we have developed to put this statement on a more quantitative footing, again examining the Gaussian case. 

@@

Note that we have assumed that prior effects are negligible (so $1\approx \infty$), and that $\epsilon \ll 1$ so that the denominator is approximately the accumulated evidence. Using \cref{eq:gaussian_logL}, we find the Gaussian termination condition in terms of lower incomplete gamma functions to be

@@

\begin{equation}\label{eq:Xf_approx}

@@

Putting this together, taking $\mathcal{L}_i$ in \cref{eq:normalised_likelihood} to be $\mathcal{L}_\mathrm{f}$ and combining this with \cref{eq:ymaxdn}, we find

@@

showing that Gaussian nested sampling chains will finish at a contour $d/2e$ away from the maximum log-likelihood. The final set of $n_i$ live points gets you $\log n_i/e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still at the $1\sigma$ interval. 

@@

Making the traditional termination criterion stricter therefore has limited returns in high dimensions, if it is ultimately still based on the remaining evidence. However, nested sampling still shrinks around the maximum exponentially, so provided a good alternative termination criterion is chosen, it will get there in reasonable time. 
To quantify this statement, consider the number of iterations $\Delta i$ required to get from the posterior mean to the true maximum, which we will now calculate. At the posterior, we have roughly speaking
    (\log X, \log\Like) = (-\DKL, \logLmax - d/2).
The prior volume $\log X_\delta$ at which the likelihood is within $\delta$ of the maximum of the Gaussian can then be found by inverting \cref{eq:gaussian_logL}:
    \log X_\delta = -\DKL - \frac{d}{2} \left(\log \frac{d}{2} - \log \delta\right),
which corresponds to an additional
\begin{equation}\label{eq:delta_i_pd}
    \Delta i = \frac{nd}{2} \left(\log \frac{d}{2} - \log \delta\right)
iterations after the bulk is reached, where $n$ is the harmonic mean of the number of live points. One would therefore expect a general distribution to take a similar order of magnitude of steps, that is $\mathcal{O}(\tfrac{n d }{2}\log \tfrac{d}{2})$ iterations to get from the usual nested sampling stopping point to within an $e$-fold of the maximum. A rule-of-thumb termination criterion could therefore be to run for at least $\tfrac{nd}{2}\log\tfrac{d}{2}$ iterations after the posterior is reached.
We can also put compare this endpoint with the termination condition based on $\epsilon$. From \cref{eq:Xf_approx}, for any $d$, $n$ we have
    \log X_\mathrm{f} \approx \frac{d}{2}\log 2\sigma^2 + \log\epsilon + \log\Gamma\left(1+\frac{d}{2}\right),
and using \cref{eq:gaussian_logL} to invert $\log\Like = \logLmax-\delta$ gives
    \log X_\delta=\frac{d}{2}\log 2\sigma^2+\frac{d}{2}\log\delta.
Putting these together means that the difference in $\log X$ between the termination criterion based on evidence fraction $\epsilon$, and a point $\delta$ away from the maximum log-likelihood is
    \log X_\delta - \log X_\mathrm{f} \approx \frac{d}{2}\log\delta - \log\epsilon - \log\Gamma\left(1+\frac{d}{2}\right).
Getting to within $\mathcal{O}(1)$ of the maximum after the termination criterion therefore takes an additional
     \Delta i \approx n \left(\log\Gamma\left(1+\frac{d}{2}\right) - \log\frac{1}{\epsilon}\right)
iterations. A larger dimension pushes the maximum away from the termination point, while a smaller $\epsilon$ brings it closer. Note the similarity of this equation to \cref{eq:delta_i_pd} via the Stirling approximation, and setting $\epsilon = 1/2$, approximately true at the posterior mean. 
It automatically follows that the $\epsilon$ one should set to end the run $\mathcal{O}(1)$ away from the maximum is simply
    \boxed{\epsilon \approx \frac{1}{\Gamma\left(1 + d/2\right)}}.
A summary of the distances between the notable points at the end of a run is shown in \cref{fig:logL_distribution}.

@@

\subsection{Temperature}\label{sec:temperature}
\subsubsection*{Motivations}
In addition to $\log \Like_i$ and $\log X_i$, another useful piece of information at each iteration is the current \textit{temperature}, which we will now define.
\par
Bayesian inference is closely related to thermodynamics, for if one equates the parameters to microstates $i$, the negative log-likelihood to the microstate energy $E_i$, and the prior to the density of states $g_i$, then the posterior as given by the generalised Bayes' rule is the canonical ensemble at inverse temperature $\beta = 1/T$:
\begin{equation}
    p(E_i) = \frac{g_i e^{-\beta E_i}}{Z(\beta)} \quad \leftrightarrow \quad \mathcal{P}_\beta(\theta) = \frac{\Like^{\beta}(\theta)\pi(\theta)}{\mathcal{Z}(\beta)}.
\end{equation}
Nested sampling therefore calculates the partition function $Z(\beta)$, with the feature that it is agnostic to the temperature; because a nested sampling chain is invariant under monotonic transformations of the likelihood, one can get the posterior at any temperature by setting $\Like \to \Like^\beta$ without losing information.
\par
Transforming the likelihood with $\beta$ changes the termination point of the run, because the posterior mass is redistributed. In the middle of a run, there is therefore a temperature one can set which makes the current iteration the termination point. This `current temperature of termination' provides a measure of progress, marching from $0 \to 1$ as the run converges, though it is uncalibrated in the sense that knowledge of the current temperature gives no indication of when the final temperature $\beta = 1$ will be reached. It is of course dependent on the threshold $\epsilon$ chosen for the termination point, and we will label this $\beta$ as $\beta_\epsilon$ to distinguish it from other definitions of the temperature we will later discuss.
\par
Changing the termination point is perhaps the most obvious consequence of changing $\beta$, but transforming the posterior mass can be useful in other ways. $\beta$ is ultimately just an additional degree of freedom in a nested sampling run, to be set at any value one sees fit. The primary purpose for introducing it in this paper is that by redistributing the posterior mass like so, one can regularize the Bayesian model dimensionality of an intermediate set of samples, which provides crucial information about the likelihood and hence the termination point. This will be elaborated on in \cref{sec:dimensionality}; for now, we discuss several other `temperatures' relevant to nested sampling, including those introduced in previous literature.

>>>>

%{{{ Preamble
\documentclass[usenatbib]{mnras}
\usepackage{graphicx}
\usepackage[export]{adjustbox}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{subcaption}
\usepackage{layouts}
\usepackage{hyperref}
\usepackage[capitalise]{cleveref}
\usepackage{fancyvrb}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage[T1]{fontenc}
%}}}

\renewcommand\thesubsubsection{\Alph{subsubsection}.}
\newcommand{\nlive}{n_i}
\newcommand{\Like}{\mathcal{L}}
\newcommand{\DKL}{\mathcal{D}_\mathrm{KL}}
\newcommand{\logLmax}{\log \Like_\mathrm{max}}

@@

\section{Introduction}
Nested sampling is a multi-purpose algorithm invented by John Skilling which simultaneously functions as a probabilistic sampler, integrator and optimiser \citep{skilling}. It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics \citep{Trotta_2008}, materials science \citep{materials} and machine learning \citep{sparse_reconstruction}. The core algorithm is unique in its estimation of volumes by \textit{counting}, coupling together nested monte carlo integrals which makes high-dimensional integration feasible and robust. It also avoids problems that challenge traditional Bayesian samplers, such as posterior multi-modality and phase transitions.
\par
The order of magnitude run-time of an algorithm, that is, whether termination is hours or weeks and months away, is of high importance to the end user. Currently, existing implementations of nested sampling (e.g. \citealt{multinest, polychord, dnest, dynesty, ultranest, nessai,proxnest}) either do not give an indication of remaining run-time, or only provide crude measures of progress that do not directly correspond to the the true endpoint.
\par
This paper sets out a principled manner of endpoint estimation for nested sampling at each intermediate stage (as shown in \cref{fig:polychord_output}), the key idea being to use the existing samples to predict the likelihood in the region we have yet to sample from. We begin with an overview of nested sampling in \cref{sec:background}, followed by an examination of the anatomy of a nested sampling run to establish key concepts for endpoint prediction in \cref{sec:anatomy}. \cref{sec:endpoint} then outlines the methodology we use, including discussion and comparisons to previous attempts. Finally, \cref{sec:results} presents the results and discussions for toy and cosmological chains, before we conclude.
\begin{figure}
\begin{Verbatim}[frame=single, commandchars=\\\{\}]
\textcolor{red}{Predicted endpoint: 25054 +/- 242}
\textcolor{red}{Progress: [=================>########] 72%}
___________________
lives      |   500 |
phantoms   | 24310 |
posteriors | 18018 |
equals     |   245 |
-------------------
ncluster   =  1/1
ndead      =  18018
nposterior =  18018
nequals    =  249
nlike      =  4159049
<nlike>    =  491.04 (9.82 per slice)
log(Z)     =  -12.55 +/- 0.27
end{Verbatim}
\caption{Output from \textsc{PolyChord} for a typical nested sampling run. The predicted endpoint, shown in red, is calculated using the method described in this paper.}
\label{fig:polychord_output}
\end{figure}

\section{Background}\label{sec:background}
We begin with a brief description of the nested sampling algorithm to establish the necessary notation. For a more comprehensive treatment, we recommend the original \citep{skilling} paper, the \citet{sivia} textbook, as well as the excellent technical review by \citet{Buchner_2023} and Nature review by \citet{physical_scientists}. 
\par
For a given likelihood $\mathcal{L}(\theta)$ and prior $\pi(\theta)$, nested sampling simultaneously calculates the Bayesian evidence
\begin{equation}\label{eq:evidence}
	\mathcal{Z} = \int \mathcal{L}\left(\theta\right)\pi(\theta)\ \mathrm{d}\theta
\end{equation}
while producing samples of the posterior distribution
\begin{equation}
	\mathcal{P}(\theta) = \frac{\mathcal{L}(\theta) \pi(\theta)}{\mathcal{Z}}.
\end{equation}
The algorithm operates by maintaining a set of $\nlive$ \textit{live points} sampled from the prior, which can vary in number throughout the run \citep{dynamic_ns}. At each iteration $i$, the point with the lowest likelihood is removed and added to a list of \textit{dead points} (illustrated in \cref{fig:dead_measure}). New points are then (optionally) drawn from the prior, subject to the constraint that they must have a higher likelihood than the latest dead point~$\mathcal{L}_i$. Repeating the procedure leads to the live points compressing around peaks in the likelihood. If there are more births than deaths, the number of live points $n_i$ increases, whilst choosing to not generate new points reduces the number of live points. The exact creation schedule depends on the dynamic nested sampling strategy.
\par
The integral in \eqref{eq:evidence} is then evaluated by transformation to a one-dimensional integral over the \textit{prior volume} $X$
\begin{equation}
	\mathcal{Z} = \int_0^1 \mathcal{L}(X)\ \mathrm{d}X \approx \sum_{i=1} \mathcal{L}_i\ \tfrac{1}{2}(X_{i-1}-X_{i+1}),
\end{equation}
where $X(\mathcal{L})$ is the fraction of the prior with a likelihood greater than $\mathcal{L}$. Whilst the likelihood contour $\mathcal{L}_i$ at each iteration is known, the prior volumes $X_i$ must be statistically estimated as follows: one can define a \textit{shrinkage factor} $t_i$ at each iteration $X_{i} = t_i X_{i-1}$, such that
\begin{equation}\label{eq:X_dist}
	X_i = \prod_{k=1}^i t_k.
\end{equation}
The $t_i$ are the maximum of $\nlive$ points drawn from $[0,1]$, so follow the distribution
\begin{equation}\label{eq:t_dist}
	P(t_i) = \nlive t_i^{\nlive-1},
\end{equation}
\begin{equation}\label{eq:t_moments}
    \langle\log t_i\rangle = -\frac{1}{\nlive}, \quad \mathrm{Var}(\log t_i) = \frac{1}{\nlive^2}.
\end{equation}
The algorithm terminates when an user-specified condition is met; a popular choice is when the evidence in the live points falls below some fraction $\epsilon$ of the accumulated evidence e.g. $10^{-3}$, which is proven to be a valid convergence criterion \citep{evans, Chopin_2010}.
Much of the existing literature treats this remaining evidence separately, for instance by estimating it as the termination $X$ multiplied by the average likelihood amongst the remaining live points. It is, however, quantitatively equivalent but qualitatively neater to consider termination as killing the remaining live points off one-by-one, incrementing the evidence exactly as during the run with decreasing $\nlive$ \citep{dynesty}.
\par
Uncertainties in the evidence are dominated by the spread in the prior volume distribution, and the simplest way to estimate them is by Monte Carlo sampling over sets of $\bm{t}$. As Skilling and others \citep{Chopin_2010, Keeton_2011} have shown, for any given problem the uncertainty in $\log \mathcal{Z}$ is proportional to $1/\sqrt{n_\mathrm{live}}$, so $n_\mathrm{live}$ sets the resolution of the algorithm.

@@

The bulk of the posterior lies within a prior volume ${X = e^{-\DKL}}$, which is the target compression. From \cref{eq:t_moments} one gets there by iteratively taking steps of size ${\Delta \log X_i = -1/n_i}$, so that when we add up the contribution of each step in \eqref{eq:t_moments} we get

@@

\caption{The dead points of a nested sampling run, recursively zoomed in. Their density is constant in $\log X$, which is a geometrically constant measure $\propto \mathrm{d} X/X$, and hence scale invariant until the final set of live points is reached. The live points at a given iteration $i$ (larger dots) are uniform across the prior, plotted for comparison.}

@@

The same is not true for the live points, which are uniformly distributed in prior volume, and by the comments in the penultimate paragraph of \cref{sec:background} have ${n_{i+k} = n_i-k}$. As a result, the maximum live point is found at 

@@

The live points therefore only get us a factor of $\log \nlive$ closer in volume to the posterior bulk. In other words, it is not until we are around $\log \nlive$ away from $\log X = -\DKL$ that the samples begin populating the posterior typical set. One can see from \eqref{eq:DKL} that the divergence increases linearly with dimension, so for large dimensionalities and typical live point numbers $\lesssim 1000$, this does not happen until near the end of the run.

@@

\subsection{Log-likelihood}\label{sec:logL}
We now consider the distribution of the samples in log-likelihood. To get an insight into the analytics, we will examine the representative case of the $d$-dimensional multivariate Gaussian with maximum point $\logLmax$ and length scale $\sigma$:
\begin{equation}\label{eq:gaussian_logL}
    \log \Like = \logLmax - X^{2/d}/2\sigma^2.
\end{equation}
Providing that $\sigma^d \ll 1$, the posterior can be approximated as 
\begin{equation}
    \mathcal{P}(X) = \tfrac{1}{(2\pi\sigma^2)^{d/2}} e^{-X^{2/d}/2\sigma^2}, 
    \label{eqn:PX}
\end{equation}
which in terms of $\log\mathcal{L}$ becomes
\begin{equation}
    \mathcal{P}(\log\mathcal{L}) = \frac{1}{\Gamma\left(\frac{d}{2}\right)}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
\end{equation}
i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, with mean and variance
\begin{equation}
    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
\end{equation}
Given the prior is uniform $X\sim U(0,1)$, the KL divergence in the $\sigma^d \ll 1$ limit is:
\begin{equation}
\mathcal{D}_\mathrm{KL} = -\frac{d}{2}\log 2 \Gamma(1+\tfrac{d}{2})^{2/d} \sigma^2.
\end{equation}

One can also derive the individual distributions for the live and dead points in log-likelihood. Note that this is merely the distribution of the \textit{values} of the points, without their nested sampling weights. The dead distribution is

@@

since the log-densities of the dead point prior volumes are proportional to the live point number at any given iteration. Meanwhile, the live points are uniformly distributed, and so have the distribution 

@@

so that $y=0$ corresponds to the latest dead point at $\log\Like_i$ and $y=1$ to the maximum. The live point distribution simplifies to

@@

which may be roughly summarised as

@@

     \pm \left( \frac{\Gamma(1+n_i)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n_i)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n_i)^2}{\Gamma(1+\frac{2}{d}+n_i)^2}\right)^{\frac{1}{2}},

@@

or in the large $d$, $n_i$ limit

@@

where $\psi^{(1)}$ is the trigamma function and $H_{n_i}$ is the $n_i$th harmonic number. This is a very small fraction in high dimensions, showing that until the end the live points are far from the posterior bulk. 

@@

    y_{(\nlive)}^{\mathrm{live}} = \frac{2\log \nlive}{d}.

@@

In other words, the highest likelihood point is always at least an order of magnitude greater than the second highest. It is therefore typically the case that nearly all of the posterior mass is concentrated in a single point, the maximum live point, until the very end of the run when the prior volumes have shrunk enough to compensate.

@@

Previous literature \citep{Akrami_2010, Feroz_2011} has explored the potential for nested sampling to be used as a global maximiser, given its ability to handle multi-modalities. In particular, the latter authors emphasised that posterior samplers such as nested sampling find the bulk of the \textit{mass}, not the maximum of the distribution, but that this can be remedied by tightening the termination criterion. We now use the machinery we have developed to put this statement on a more quantitative footing. 

@@

Note that we have assumed that prior effects are negligible (so $1\approx \infty$), and that $\epsilon \ll 1$ so that the denominator is approximately the accumulated evidence. Computing this for \eqref{eq:gaussian_logL}, we find the answer in terms of lower incomplete gamma functions

@@

@@

Putting this together, taking $\mathcal{L}_i$ in \eqref{eq:normalised_likelihood} to be $\mathcal{L}_\mathrm{f}$ and combining this with \eqref{eq:ymaxdn}, we find

@@

showing that in general nested sampling will finish at a contour $d/2e$ away from the maximum log-likelihood. The final set of $n_i$ live points gets you $\log n_i/e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation. 

@@

Making the traditional termination criterion stricter therefore has limited returns in high-dimensions, if it is ultimately still based on the remaining evidence. However, nested sampling still shrinks around the maximum exponentially, so provided a good alternative termination criterion is chosen, it will get there in reasonable time.

To quantify this statement, consider the number of iterations $\Delta i$ required to get from the posterior bulk to the true maximum, which we will now calculate. At the posterior, we have roughly speaking
\begin{equation}
    (\log X, \log\Like) = (-\DKL, \logLmax - d/2).
\end{equation}
The prior volume $\log X_\delta$ at which the likelihood is within $\delta$ of the maximum can then be found by inverting \cref{eq:gaussian_logL}:
\begin{equation}
    \log X_\delta = -\DKL - \frac{d}{2} \left(\log \frac{d}{2} - \log \delta\right),
\end{equation}
which corresponds to an additional
\begin{equation}
    \Delta i = \frac{nd}{2} \left(\log \frac{d}{2} - \log \delta\right)
\end{equation}
iterations after the bulk is reached, where $n$ is the harmonic mean of the number of live points. One would therefore expect a general distribution to take $\mathcal{O}(\tfrac{n d }{2}\log \tfrac{d}{2})$ iterations to get from the usual nested sampling stopping point to within an $e$-fold of the maximum. A rule-of-thumb termination criterion could therefore be to run for at least $\tfrac{nd}{2}\log\tfrac{d}{2}$ iterations after the posterior is reached.
\par
A summary of the distances between the notable points at the end of a run is shown in \cref{fig:logL_distribution}.

@@

\subsection{Temperature}\label{sec:temperature}
\subsubsection*{Motivations}
As shown in the previous section, midway through the run nearly all of the posterior mass is concentrated at a single point. However, this does not capture the \textit{structure} of the posterior that has been explored and all of the information it provides. 
\par
We have the potential to fix this because nested sampling is invariant to monotonic transformations, so we can transform the likelihood as $\Like \to \Like^{\beta}$ without loss of information by trivially re-weighting the samples. Increasing $\beta$ worsens the situation, while $\beta \to 0$ simply gives back the prior. There is, on the other hand, a significant intermediate range which makes the samples look like a posterior centred at the present contour, which will allow us to recover the structure of the samples. A schematic of the procedure is shown in \cref{fig:last_live_point}.
\par
\begin{figure}
\begin{center}
    \includegraphics{figures/last_live_point.pdf}
\end{center}
\caption{The likelihood and posterior as a function of $\log X$ in the middle of a nested sampling run. Almost all of the posterior mass is concentrated at a single point with the highest compression, because it is orders of magnitude higher in likelihood. Reducing $\beta$ re-weights to shift the posterior mass; taking $\beta \to 0$ goes too far and gives back the prior, but there is some intermediate $\beta = \beta_i$ which is significant because it centres the posterior at the current contour.}
\label{fig:last_live_point}
\end{figure}
\par
At this point it is relevant to note the correspondence between Bayesian inference and statistical mechanics, from which the above transform is derived. If one equates the parameters to microstates $i$, the negative log-likelihood to the microstate energy $E_i$, and the prior to the density of states $g_i$, then the posterior as given by the generalised Bayes' rule is the canonical ensemble
\begin{equation}
    p(E_i) = \frac{g_i e^{-\beta E_i}}{Z(\beta)} \quad \leftrightarrow \quad \mathcal{P}_\beta(\theta) = \frac{\mathcal{L}^{\beta}(\theta)\pi(\theta)}{\mathcal{Z}(\beta)}
\end{equation}
at the inverse temperature $\beta = 1/T$. As noted by \citet{demons}, thermal algorithms such as thermodynamic integration \citep{path_sampling} get the evidence by evolving through a series of canonical ensembles via some temperature schedule, but nested sampling instead maintains \textit{microcanonical} ensembles, which are the iso-likelihood contours. Instead of using temperature \citep{simulated_annealing, parallel_tempering} or energy \citep{wang_landau} as a control parameter, nested sampling chooses a series of ensembles with constant relative volume entropy $\Delta \log X$, which allows the algorithm to handle phase transitions \citep{baldock}.
\par
Because the temperature of a microcanonical ensemble is a derived property rather than a parameter, there is some freedom in its definition. Returning to our original motivation, we make the connection that the temperature is the re-weighting $\Like \to \Like^{\beta}$ which centres the ensemble around the current energy. We now present several temperatures that achieve this aim, each of which one can plausibly consider to be the current temperature of a nested sampling run.
